Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-09 paper/ms.tex
update ms

Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
help in building the timeline for exoplanet evolution.  Informative
individual young systems include V1298~Tau, an adolescent resonant
chain of four transiting puffy planets that is a likely precusor to
the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
Thao et al.~submitted).  Population-level analyses have similarly
suggested differences in the size distribution of young exoplanets
relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, or of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has matured over the past decade, primarily due to Kepler, K2, TESS,
and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

@@

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk?  Second, where are the missing young transiting planets?
Briefly, we will find that uniform is incorrect by a factor of a few,
and that at least one bias against the discovery of young transiting
planets is in the measurement of their ages.

@@

However, beyond improvements in rotation-based ages, recent years have
also yielded improvements in the lithium age scale.  Lithium ages
include two qualitatively distinct regimes: depletion boundary ages
for M-dwarfs in star clusters, and the less precise decline-based ages
for individual field FGK stars \citep{Soderblom_2010}.  Here we focus
on the latter approach, which relies on the observation that the Li
abundances of partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who built a model for how the equivalent width
(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
stellar effective temperature and age for a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our methods for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

A few new highlights from the high quality stars and planets include
Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.
Kepler-1529b is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured
rotation and lithium ages.
Kepler-1565b is an analogous $\approx$170-230\,Myr super-Earth.
Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr near-Solar analog with
two known transiting planets: an Earth-sized planet on a one-day orbit,
and a mini-Neptune on a five-day orbit.
The Earth-sized planet, Kepler-1312c, along with the near-identical
Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), 
could be the
youngest Earth-sized planets currently known, comparable to systems
such as HD~63433d \citep[1.1\,$R_\oplus$,
$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b \citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
Kepler-1629~b, with a size just two-thirds that of Earth
(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on rotation
(\kepsixteentwonine\,Myr).  The lithium age from a reconnaissance 
TRES spectrum confirms the point.

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates of main sequence stars.
Binarity is one example: even intermediate-separation ($\sim$10\,AU) binaries are biased
toward rapid rotation
\citep[e.g.][and many studies thereafter]{Meibom_2007}. %,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
In this work, we were interested in not only the overall
age distribution of the Kepler field, but also in the reliability of individual ages for
young planet hosting stars.
The quality flagging scheme in Table~\ref{tab:planets} provides one path toward this end.
The other method is to consider the region in which we do have
overlapping sensitivity between lithium and rotation -- 
G and K dwarfs with $t$$\lesssim$1\,Gyr.  
Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in agreement
(with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a clearly
constraining lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

{\it Field star ages from other surveys}:
In this study we focused on deriving rotation and lithium-based ages for the Kepler
planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
Rotation-based age dating of their discoveries could be useful for the effort
of identifying young exoplanets, and also for the broader issue of understanding
their evolution.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
connected to their birth clusters.
Others have not.
A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
yet to be performed.
The ideal approach for such a search would be to combine spatial and kinematic information
from Gaia with rotation measurements from TESS.
Given that the decoherence time of young clusters is $\sim$100\,Myr, one would expect that
the youngest ``field stars'' may have young neighbors that are relatively easy to
identify.

>>>>

Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
help build the timeline for exoplanet evolution.  Informative
individual young systems include V1298~Tau, an adolescent resonant
chain of four transiting puffy planets that is a likely precusor to
the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
Thao et al.~submitted).  Population-level analyses have similarly
suggested differences in the size distribution of young exoplanets
relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, or of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the galactic star
formation rate.

Despite this and other observational challenges, young close-in planet discovery
has matured over the past decade, primarily due to Kepler, K2, TESS,
and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets is often easier than
finding the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

@@

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for transiting planet host star ?
Second, where are the missing young planets?  Briefly, we will find
that uniform is incorrect by a factor of a few, and that at least one
bias against the discovery of young transiting planets is in the
measurement of their ages.

@@

However, beyond rotation-based ages, recent years have
also yielded improvements in the lithium age scale.  Lithium ages
include two distinct regimes: depletion boundary ages
for M-dwarfs in star clusters, and the less precise decline-based ages
for individual field FGK stars \citep{Soderblom_2010}.  Here we focus
on the latter approach, which relies on the observation that the Li
abundances of partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}. 
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who built a model for how the equivalent width
(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
stellar effective temperature and age for a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K dwarfs lose their surface lithium much
faster than G dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our method for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

A few new highlights from the high quality stars and planets include
Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.
Kepler-1529b is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured
rotation and lithium ages.
Kepler-1565b is an analogous $\approx$170-230\,Myr super-Earth.
Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr near-Solar analog with
two transiting planets: an Earth-sized planet on a one-day orbit,
and a mini-Neptune on a five-day orbit.
The Earth-sized planet, Kepler-1312c, along with the near-identical
Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), 
rank among the
youngest Earth-sized planets currently known, comparable to systems
such as HD~63433d \citep[1.1\,$R_\oplus$,
$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b \citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
Kepler-1629~b, with a size just two-thirds that of Earth
(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on rotation
(\kepsixteentwonine\,Myr).  The lithium age from a reconnaissance 
TRES spectrum confirms the point.

@@

Most of our ages come from one source of information: rotation.
Factors other than stellar age and mass can influence rotation rates of main sequence stars.
Binarity is one example: even intermediate-separation ($\sim$10\,AU) binaries are biased
toward rapid rotation
\citep[e.g.][and many studies thereafter]{Meibom_2007}. %,2016AJ....152..115S,2019ApJ...871..174S,Gillen_2020,Bouma_2021}.
Metallicity may also be a second-order effect
\citep{2020MNRAS.499.3481A,2024arXiv240500779S}.
In this work, we were interested in not only the overall
age distribution of the Kepler field, but also in the reliability of individual ages for
young planet hosting stars.
The quality flagging scheme in Table~\ref{tab:planets} provides one path toward this end.
The other method is to consider the region in which we do have
overlapping sensitivity between lithium and rotation -- 
G and K dwarfs with $t$$\lesssim$1\,Gyr.  
Our results for such stars include
\ltonegyrhighqconfirmedtwosided\ planets orbiting ``high-quality''
stars for which the two age indicators are both constraining and in agreement
(with $Q_{\rm star}$ bits zero through nine not raised, and no planet quality flags raised).
Conversely, we found only one system, Kepler-786, with a 
precise lithium age (\kepseveneightsix\,Myr), but no rotation signal.

@@

{\it Field star ages from other surveys}:
In this study we focused on deriving rotation and lithium-based ages for the Kepler
planets.  However surveys such as K2, TESS, HAT, and WASP have had very different selection functions.
Rotation-based age dating of their discoveries could 
potentially clarify their youth, and thereby help clarify the broader
issue of their evolution.

{\it Searches for new associations}:
Some of the youngest systems in Table~\ref{tab:planets} have been
connected to their birth clusters.
Others have not.
A systematic search for the birth clusters of the youngest ``field stars'' has, to our knowledge,
not been performed.
The ideal approach for such a search would be to combine spatial and kinematic information
from Gaia with rotation measurements from TESS.
Given that the decoherence time of young clusters is $\sim$100\,Myr, one would expect that
the youngest ``field stars'' may have young neighbors that are relatively easy to
identify.
